---
title: "HW3"
author: "Jackson Dial"
date: '2023-02-08'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(pander)
library(lubridate)
```

# Data Curation/Data Mining Tasks

## Task 1

I used the CLI through gitbash to extract the tar file using the command:

_tar -xvzf diabetes-data.tar.Z -C ._

Once the data files were extracted, I used python to loop through each of the data files and concatenating them together into one data file. See attached file "combine_data.py" file for code.


## Task 2

```{r}
dat <- read_csv("combined_data.csv")

dat1 <- dat %>% 
  select(2:5) %>% 
  rename("date" = "0", "time" = "1", "code" = "2", "value" = "3") %>% 
  separate(date, sep = "-", into = c("month", "day", "year"), remove = FALSE) %>% 
  separate(time, sep = ":", into = c("hour", "minute"), remove = FALSE) %>% 
  arrange(year, month, day, hour, minute) %>%
  select(date, time, code, value)
```

```{r}
sum(is.na(dat1$date));sum(is.na(dat1$time));sum(is.na(dat1$code));sum(is.na(dat1$value));sum(is.na(dat1$day_of_week))
```




__Submit this dataset^^__

## Task 3

```{r}
nrow(dat1)
```
The total number of measurements in the dataset is 29330.

## Task 4

```{r}
year_dat <- dat1 %>% separate(date, sep = "-", into = c("month", "day", "year"))
year_dat %>% count(year) %>% pander()
```


## Task 5


```{r}
dat1$day_of_week <- wday(mdy(dat1$date), label = TRUE, abbr = FALSE)
dat1 %>% count(day_of_week) %>% pander()
```


## Task 6

```{r}
dat1 %>% count(code) %>% pander()
```

## Task 7

There are a few missing values in the data.

```{r}
sum(is.na(dat1))
```
There are 106 missing values (cells) in the dataset. 

```{r}
sum(!complete.cases(dat1))
```
But only 73 rows with missing values.

## Task 8

I think the impact of missing or incorrect data on visualizations can be easily overlooked. If there is an incorrect value that becomes an outlier, then it is typically easy to notice in a visualization. But, if there are missing values, they will typically not be presented at all in the visualization and if warnings in the code are ignored, it will be as if they do not exist. It is important to note missing values, but it is not always necessary to remove or to impute them.

Invalid or incomplete data will have a negative impact on faceted search in that if a certain facet specification is missing, that item (or observation) will be excluded from the query results.

# Data Visualization Tasks

## Task 1

```{r}
#change value to be a numeric variable
dat1$value <- as.numeric(dat1$value)

ggplot(dat1, aes(x = value))+
  geom_histogram(stat = "count",fill = "dodgerblue")+
  theme(panel.grid.minor = element_blank())
```

I needed to change the variable type of the value column to be numeric instead of a character. The plotting package also automatically removes missing values, so I did not need to remove them from the actual dataset.

## Task 2

```{r}
dat1 %>% 
  group_by(day_of_week) %>% 
  summarise(min = min(value, na.rm = TRUE), Q1 = quantile(value, .25, na.rm = TRUE), mean = mean(value, na.rm = TRUE), median = median(value, na.rm = TRUE), Q3 = quantile(value, .75, na.rm = TRUE), max = max(value, na.rm = TRUE)) %>% pander()

ggplot(dat1, aes(x = day_of_week, y = value))+
  geom_boxplot(aes(fill = day_of_week))+
  stat_summary(fun = "mean", color = "red", shape = 4)+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 30, vjust = .8),
        axis.ticks = element_blank(),
        plot.subtitle = element_text(color = "red"))+
  labs(x = "Day of the Week",
       y = "Blood Glucose Level",
       fill = "Day of the Week",
       title = "Blood Glucose Levels by Day of the Week",
       subtitle = "Red X Denotes Mean", color = "red")
```

The data did need to be pre-processed a bit, but I did it inline with the table code in order to simplify my code and to not edit the originally read-in dataset. This pre-processing that was needed was to not include the NA or missing values in numeric calculations for the mean and 5-number summary.

There is a slight difference in distribution by day, but it does not seem to be very significant from a visualization perspective. The slight trend that I am observing is the higher 3rd quartile for Saturday and Sunday. This is potentially related to these days being the weekend, and people tend to eat less healthy on the weekends, which I further hypothesize is related to blood glucose levels.

A simple experimental design to allow me to test this hypothesis would be to group all variables by weekday or weekend (Saturday and Sunday), and run a simple t-test between means of the groups.


## Task 3

```{r}
ggplot(dat1,aes(x = code))+
  geom_density(aes(fill = day_of_week))+
  facet_wrap(~day_of_week, ncol = 2)+
  theme(panel.grid.minor = element_blank(),
        axis.ticks = element_blank())+
  labs(fill = "Day of Week",
       x = "Day of Week",
       y = "Density",
       title = "Faceted Distributions of Test Codes by Day of the Week")
```

This plot shows that the distributions of test codes is approximately equal by day.

## Task 4

```{r}
top5 <- dat1 %>% count(code, sort = TRUE) %>% slice(1:5) %>% select(code)

pTop5 <- dat1 %>% filter(code %in% top5$code) %>% 
ggplot(aes(x = day_of_week, y = value))+
  geom_boxplot(aes(fill = day_of_week))+
  stat_summary(fun = "mean", color = "red", shape = 4)+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 30, vjust = .8),
        axis.ticks = element_blank(),
        plot.subtitle = element_text(color = "red"))+
  labs(x = "Day of the Week",
       y = "Blood Glucose Level",
       fill = "Day of the Week",
       title = "Blood Glucose Levels by Day of the Week (Top 5 Codes Only)",
       subtitle = "Red X Denotes Mean", color = "red")


pNotTop5 <- dat1 %>% filter(!code %in% top5$code) %>% 
ggplot(aes(x = day_of_week, y = value))+
  geom_boxplot(aes(fill = day_of_week))+
  stat_summary(fun = "mean", color = "red", shape = 4)+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 30, vjust = .8),
        axis.ticks = element_blank(),
        plot.subtitle = element_text(color = "red"))+
  labs(x = "Day of the Week",
       y = "Blood Glucose Level",
       fill = "Day of the Week",
       title = "Blood Glucose Levels by Day of the Week (Top 5 Codes Excluded)",
       subtitle = "Red X Denotes Mean", color = "red")

pTop5
pNotTop5

```

The distributions of blood glucose levels does appear to be slightly different between the top 5 codes and the not-top 5 codes. I would not change my hypothesis for either of the previous tasks. Though it appears Wednesday has a higher 3rd quartile of blood glucose level in the not-top 5 code plot, it does not seem to be significantly different from the rest of the weekdays.
